The brightness and spatial distributions of terrestrial radio sources

### 2013-01-24 articles/Article-RFI-distribution/article-rfi-distribution.tex
Leons and others comments

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{3,4}$ and
S.~Zaroubi$^3$%,
%W.~Brouw$^{3,4}$,
%B.~Ciardi$^5$,
%I.~T.~Iliev$^6$,
%\newauthor
%G.~J.~A.~Harker$^7$,
%L.~V.~E.~Koopmans$^3$,
%G.~Mellema$^8$,
%S.~J.~Wijnholds$^4$
\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
$^{3}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{4}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
%$^{5}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
%$^{6}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
%$^{7}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
%$^8$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by estimating the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$ measured in flux density. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with (amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{\zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. A perfect RFI detector would separate the samples in two distributions; one that is not affected by RFI, and therefore contains noise only, and one that is the sum of RFI and noise. Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless. Even more, we still need to take into account that the samples classified as RFI are also affected by noise. Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.

@@

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. It is likely that most RFI sources are observed at the edges of the beam. Because most sources will be observed with similar gains (within one order of magnitude), it can be expected that the beam will have a limited effect on the histogram properties of an observation, and is comparable with the effect of the frequency response.

@@

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\subsection{Error analysis}
An estimate for the standard deviation of the slope estimator $\hat \alpha$ can be found by calculating $\textrm{SE}(\hat \alpha)$, the \emph{standard error} of $\hat \alpha$. The standard error of the slope of a straight line \citep[pp. 32--35]{acton-analaysis-of-straight-lines} is given by
\begin{equation} \label{eq:stderr-slope}
 \textrm{SE}(\hat \alpha) = \sqrt{\frac{SS_{yy}-\hat\alpha SS_{xy}}{\left(n - 2\right) SS_{xx}}},
\end{equation}
where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples. However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed. The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples. Consequently, the estimated standard deviation of $\hat \alpha$ is very low. Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error. Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range. This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$. If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller. Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
\begin{equation}
 \epsilon_{\hat\alpha} = \sqrt{\frac{\sum \left( \alpha_i - \hat\alpha \right)^2 }{n^2_\alpha - n_\alpha}}.
\end{equation}
This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.

@@

\subsection{Histogram analysis}
Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets. In both sets, it is clear that at least one component with a Rayleigh and one with a power-law distribution have been observed. The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution. The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response as discussed in \S\ref{sec:parameter-variability}. As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels. This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ confidence boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

On the assumption that the power-law distribution for RFI sources will continue down into the noise, we have constructed a full parametrization of the RFI apparent flux distribution. By assuming that all samples contain some contribution of RFI, we find that the average flux density of RFI that leaks through the detector is $484$--$496$~mJy for the LBA and $167$--$171$~mJy for the HBA (depending on the used method). These values should be compared to the noise in individual samples of $770$~Jy (LBA) and $77$~Jy (HBA) (see Table~\ref{table:dist-data-quantities}), and are upper limits for what can be expected. If in fact not all samples are affected by RFI, the leaked RFI flux will be smaller, and will of course be zero in the extreme case that the detector has found and removed all RFI.

In experiments such as the LOFAR EoR project, a simulation pipeline is used to create a realistic estimate of the signal that can be expected. Currently, these simulations do not include the effects of RFI. With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation. Using Eq.~\eqref{eq:sample-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger. What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data. For example, it is neither realistic that all RFI sources are continuously transmitting nor that they affect only one sample. The RFI detector is highly depending on the morphology of the feature in the time-frequency domain. Finally, the coherency properties of the RFI might be even more important to simulate correctly, but these have been not been explored. However, these have large implications for observations with high sensitivity. This will be discussed in the next section.

@@

On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.

>>>>

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{3,4}$,
S.~Zaroubi$^3$,
L.~V.~E.~Koopmans$^3$,
\newauthor
S.~J.~Wijnholds$^4$,
F.~B.~Abdalla$^5$,
W.~Brouw$^{3,4}$,
B.~Ciardi$^6$,
I.~T.~Iliev$^7$,
G.~J.~A.~Harker$^8$,
\newauthor
G.~Mellema$^8$\\
$^1$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^2$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
$^3$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^4$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by estimating the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$ measured in flux density. This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects. The antennas are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. The formula represents a spherically propagating wave in free space. We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{\zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram. According to Hata's model, this is valid for the range 30--200 m. The trend of the slope will not continue indefinitely when increasing the height further. Instead it will converge to a $-2$ slope, which corresponds with free-space propagation.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one one group of samples that are contaminated, and are the sum of RFI and noise. Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless. Even more, we still need to take into account that the samples classified as RFI are also affected by noise. Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.

@@

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized near the horizon. It is likely that most RFI sources are observed at the edges of the beam. Nevertheless, most sources will be observed with similar gains (within one order of magnitude), and it can be expected that the beam will have a limited effect on the histogram properties of an observation. It is therefore comparable with the effect of the frequency response.

@@

$S_1$ and $S_2$ are selected by visually inspection of the histogram. Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the average flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of data noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\subsection{Error analysis}
An estimate for the standard deviation of the slope estimator $\hat \alpha$ can be found by calculating $\textrm{SE}(\hat \alpha)$, the \emph{standard error} of $\hat \alpha$. The standard error of the slope of a straight line \citep[pp. 32--35]{acton-analaysis-of-straight-lines} is given by
\begin{equation} \label{eq:stderr-slope}
 \textrm{SE}(\hat \alpha) = \sqrt{\frac{SS_{yy}-\hat\alpha SS_{xy}}{\left(n - 2\right) SS_{xx}}},
\end{equation}
where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples. However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed. The noise in the part of the histogram over which the slope is calculated, is very small due to the large number of samples. Consequently, the estimated standard deviation of $\hat \alpha$ is very low. Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error. Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range. This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$. If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller. Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
\begin{equation}
 \epsilon_{\hat\alpha} = \sqrt{\frac{\sum \left( \alpha_i - \hat\alpha \right)^2 }{n^2_\alpha - n_\alpha}}.
\end{equation}
This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.

@@

\subsection{Histogram analysis}
Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets. In both sets, it is clear that at least one component with a Rayleigh and one with a power-law distribution have been observed. The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution. The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response, i.e., variability of $g$ as discussed in \S\ref{sec:parameter-variability}. As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels. This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

On the assumption that the power-law distribution for RFI sources will continue down into the noise, we have constructed a full parametrization of the RFI apparent flux distribution. By assuming that all samples contain some contribution of RFI, we find that the average flux density of RFI after excision by automated flagging is $484$--$496$~mJy for the LBA and $167$--$171$~mJy for the HBA. These values should be compared to the noise in individual samples of $770$~Jy (LBA) and $77$~Jy (HBA) (see Table~\ref{table:dist-data-quantities}), and are upper limits for what can be expected. If in fact not all samples are affected by RFI, the leaked RFI flux will be smaller, and will of course be zero in the extreme case that the detector has found and removed all RFI.

In projects such as the EoR detection experiment with LOFAR, a simulation pipeline is used to create a realistic estimate of the signal that can be expected. Currently, these simulations do not include the effects of RFI. With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation. Using Eq.~\eqref{eq:sample-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger. What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data. For example, it is neither realistic that all RFI sources are continuously transmitting nor that they affect only one sample. The RFI detector is highly depending on the morphology of the feature in the time-frequency domain. Finally, the coherency properties of the RFI might be even more important to simulate correctly, but these have been not been explored. However, these have large implications for observations with high sensitivity. This will be discussed in the next section.

@@

On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation. If, say, only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
